### 2012-11-15 thesis/MA-Thesis.tex
final!

\vspace{1cm}
\begin{flushleft}
%select german for formatting the date
\selectlanguage{ngerman}
Augsburg, den \today \hfill Ludwig Nägele
\end{flushleft}

@@

For my master thesis it was my wish to do research abroad in a foreign country. Professor Wolfgang Reif supported me in my plan and managed to give me the unique possibility to visit New Zealand for half an year. Therefore I am deeply grateful to him.

My thesis project was hostet at the University of Auckland, New Zealand, by Professor Bruce MacDonald. I want to thank him for the opportunity to stay and to work at his lab, for all the meetings and discussions we had and for all the support he gave to me.

@@

Sincere thanks is also given to Alwin Hoffmann and my supervising tutor Andreas Angerer for assisting me by discussions and hints. They did not only inspect and correct the thesis writing regularly, but they also assisted me in finding the thesis topic in the very beginning and contributed to the results of this work decisively by providing ideas and feedback during the research phase.

@@

\pdfbookmark[1]{Inhaltsverzeichnis}{toc}
\tableofcontents
\newpage
\listoffigures
\newpage
%\listoftables
\lstlistoflistings
%\pagenumbering{arabic}

>>>>

\vspace{1cm}
\begin{flushleft}
%select german for formatting the date
\selectlanguage{ngerman}
Augsburg, den 16. November 2012 \hfill Ludwig Nägele %\today
\end{flushleft}

@@

For my master thesis it was my wish to do research abroad in a foreign country. Professor Wolfgang Reif, supervisor of my thesis, supported me in my plan and managed to give me the unique possibility to visit New Zealand for half an year. Therefore I am deeply grateful to him.

My thesis project was hostet at the University of Auckland, New Zealand, by Professor Bruce MacDonald. I want to thank him for the opportunity to stay and to work at his robotics lab, for all the meetings and discussions we had and for all the support he gave to me.

@@

Sincere thanks is also given to Alwin Hoffmann and my supervising tutor Andreas Angerer, both doctoral candidates at the University of Augsburg, Germany, for assisting me by discussions and hints. They did not only inspect and correct the thesis writing regularly, but they also assisted me in finding the thesis topic in the very beginning and contributed to the results of this work decisively by providing ideas and feedback during the research phase.

@@

\pdfbookmark[1]{Contents}{toc}
\tableofcontents
\newpage
\listoffigures
\newpage
%\listoftables
\lstlistoflistings
%\pagenumbering{arabic}

### 2012-11-15 thesis/tex/2_background.tex
final!

%Zugriff auf devices via web services
%Windows is used as operating system on the robots and the programs are executed in Adobe flash and Actionscript. External devices are accessed over web

@@

In the beginning all programming of the robot behaviour was done by directly editing the XML code mentioned in the previous section by using a text editor. However, huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise. For this reason, Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL. This environment- it is called Robostudio~\cite{robostudio} - allows to edit easily the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically. This file can be downloaded to the robot and executed by the interpreter. Figure~\ref{fig:robostudio_proceeding} illustrates the steps of robot behaviour development.

@@

Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application~\cite{netbeans}.
Several windows and views provide visual tools for editing and understanding robot behaviour as well as screen dialog design. As depicted in figure~\ref{fig:robostudio_vpe}, the editor integrates a \emph{State Navigator Window} (1) in the top left corner where all existing states are listed. Below an overview about all incoming and outgoing transitions is given. New states can be created and existing states can be deleted. A click on one particular state id in the \emph{State Navigator Window} will cause all other components to show all possible information regarding the selected state.
The \emph{Expected Events Editor Window} (2) in the bottom left can be used for specifying the expected events and their effects.
Heart and center of the editor is the \emph{UI Component Layout Editor Window} (3). It renders the screen dialog of the selected state and gives a preview of the screen presented on the robot during runtime. Components such as buttons, text boxes or images can be added, changed or removed via drag and drop.
They are provided by the \emph{UI Components Palette Window} (4) on the right side where the developer can choose from all supported UI components.
The \emph{Background Actions Editor Window} (5) lets the developer define background actions, which are executed transparently to the developer such as accessing external devices or sending messages.
Aiming to support the developer with helpful graphical tools, the \emph{State Transition Visualization Window} (6) gives an overview about all connections towards and from the selected state. It helps to get quickly an understanding of the local program behaviour.

@@

The use and application of software is versatile and varies from text editing programs running on personal computers, over control software for high-precision applications in aeronautics, right up to trigger mechanisms for deathly weapons. For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also to human lifes. In order to eliminate software malfunction, there are different mathematical concepts for verifying software behaviour.
One of them is model checking which allows checking state machine programs for functional safety. Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realised with state machines, the concept of behaviour checking is applicable and important for this kind of programs. At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an introduction into model checking in section~\ref{sec:modelcheckingandltl}.

@@

Each system status is represented by one \emph{state} which stands for a particular and unique constellation of system variables and properties. One state can be declared as \emph{start state} which is the very first state when the system starts working. The state representing the current system status is called \emph{current state}.
The current state changes (maybe into itself) whenever there is a modification to the system. Such transfers of the system from the current state to another one are defined by \emph{transitions}. Thus, they define the possible system behaviour.
All influences of the environment which cause a change to the system are called \emph{events}. They trigger transitions which transfer the system into a new (current) state. The fact that a state machine reacts to events is the reason why state machines are also called \emph{reactive systems}.

@@

The short description given in this section does not contain a complete explanation about the concept of state machines at all, but it gives a basic overview which is needed as background information for this work.
Also the explanation of state machines above is no formal definition; it can be read found in literature~\cite{statemachines, statemachinesandformalmethods}.
%Also the formal definition of state machines is not relevant and can be found in literature~\cite{statemachines, statemachinesandformalmethods}.

%mathematical model of computation
%finite (endliche anzahl zustände)

@@

\begin{description}
	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true at least until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true caused by the first occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
	\item[$\varphi \mathcal{W} \psi$] ``WEAK UNTIL'' represents the same as ``UNTIL'', except that an occurrence of $\psi$	is not necessary. In this case, $\varphi$ remains true infinitely.
\end{description}

@@

Even though this approach touches the subject of healthcare it barely matches the problem stated in this work. HomeTL focuses more on monitoring temporal boundaries of a patient's behaviour than on ensuring functional safety of an implemented program. It does however depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.

>>>>

@@

In the beginning all programming of the robot behaviour was done by directly editing the RBDL code mentioned in the previous section by using a text editor. However, huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise. For this reason, Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL. This environment - it is called Robostudio~\cite{robostudio} - allows to easily edit the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically. This file can be downloaded to the robot and executed by the interpreter. Figure~\ref{fig:robostudio_proceeding} illustrates the steps of robot behaviour development.

@@

Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application~\cite{netbeans}.
Several windows and views provide visual tools for editing and understanding robot behaviour as well as screen dialog design. As depicted in figure~\ref{fig:robostudio_vpe}, the editor integrates a \emph{State Navigator Window} (1) in the top left corner where all existing states are listed. Below an overview about all incoming and outgoing transitions is given. New states can be created and existing states can be deleted. A click on one particular state id in the \emph{State Navigator Window} will cause all other components to show all possible information regarding the selected state.
The \emph{Expected Events Editor Window} (2) in the bottom left can be used for specifying the expected events and their effects.
Heart and center of the editor is the \emph{UI Component Layout Editor Window} (3). It renders the screen dialog of the selected state and gives a preview of the screen presented on the robot during runtime. Components such as buttons, text boxes or images can be added, changed or removed via drag and drop.
They are provided by the \emph{UI Components Palette Window} (4) on the right side where the developer can choose from all supported UI components.
The \emph{Background Actions Editor Window} (5) lets the developer define background actions, which are executed transparently such as accessing external devices or sending messages.
Aiming to support the developer with helpful graphical tools, the \emph{State Transition Visualization Window} (6) gives an overview about all connections towards and from the selected state. It helps to quickly get an understanding of the local program behaviour.

@@

The use and application of software is versatile and varies from text editing programs running on personal computers, over control software for high-precision applications in aeronautics, right up to trigger mechanisms for deathly weapons. For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also to human lifes. In order to eliminate software malfunction, there are different mathematical concepts for verifying software behaviour.
One of them is model checking which allows checking state machine programs for functional safety. Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realized with state machines, the concept of behaviour checking is applicable and important for this kind of programs. At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an introduction into model checking in section~\ref{sec:modelcheckingandltl}.

@@

Each system status is represented by one \emph{state} which stands for a particular and unique constellation of system variables and properties. One state can be declared as \emph{start state} which is the very first state when the system starts working. The state representing the current system status is called \emph{current state}.
The current state changes (maybe into itself) whenever there is a modification to the system. Such transfers of the system from the current state to another one are defined by \emph{transitions}. Thus, they define the possible system behaviour.
All influences of the environment which cause a change to the system are called \emph{events}. They trigger transitions which transfer the system into a new (current) state. The fact that a state machine reacts to events is the reason why state machines are also known as \emph{reactive systems}.

@@

The short description given in this section does not contain a complete explanation about the concept of state machines at all, but it gives a basic overview which is needed as background information for this work.
Also the explanation of state machines above is no formal definition; it can be found in literature~\cite{statemachines, statemachinesandformalmethods}.

@@

\begin{description}
	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true at least until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true at least as long as there is no occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
	\item[$\varphi \mathcal{W} \psi$] ``WEAK UNTIL'' represents the same as ``UNTIL'', except that an occurrence of $\psi$	is not necessary. In this case, $\varphi$ remains true infinitely.
\end{description}

@@

Even though this approach touches the subject of healthcare it barely matches the problem stated in this work. HomeTL focuses more on monitoring temporal boundaries of a patient's behaviour than on ensuring functional safety of an implemented program. It does however depict how nontechnical constraint design can be realized and its graphical design is a good example of handling time dependencies.

### 2012-11-15 thesis/tex/3_goals.tex
final!

Furthermore some mechanism should be integrated aiming to support application developers in finding suitable constraints for a certain application. It is assumed that applications are modeled by state transition graphs. This modeling paradigm is used in many healthcare robots that employ dialogue systems for communicating with robot users~\cite{Bickmore2006}. Other possible fields might be shop assistant robots or entertainment robots. For such systems, the approach is suitable. It is based on heuristically analyzing state transition graphs and deriving constraints related to the graph structure. This work investigates a particular heuristic as a first step towards developer support.

@@

Some of the statements do not make sense if patients lie or pretend to take their medication but don't take it. As the presented approach validates state machine behaviour rather than human attitudes, it is assumed that there is either cooperation from the patient or a separate tool that can verify medication intake. Furthermore the correctness of external services used by the medication reminder such as the database module holding all reminders is not taken into account.

>>>>

Furthermore some mechanisms should be integrated aiming to support application developers in finding suitable constraints for a certain application. It is assumed that applications are modeled by state transition graphs. This modeling paradigm is used in many healthcare robots that employ dialogue systems for communicating with robot users~\cite{Bickmore2006}. Other possible fields might be shop assistant robots or entertainment robots. For such systems, the approach is suitable. It is based on heuristically analyzing state transition graphs and deriving constraints related to the graph structure. This work investigates a particular heuristic as a first step towards developer support.

@@

Some of the statements do not make sense if patients lie or pretend to take their medication but do not take it. As the presented approach validates state machine behaviour rather than human attitudes, it is assumed that there is either cooperation from the patient or a separate tool that can verify medication intake. Furthermore the correctness of external services used by the medication reminder such as the database module holding all reminders is not taken into account.

### 2012-11-15 thesis/tex/5_automatedgenerationofsafetyconstraints.tex
final!

Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications. In order to discover whether a program meets certain safety requirements, its behaviour is checked by the constraints. This happens mostly as a last check before software gets released. 
But regarding updating and maintenace work, often further changes are made to the software even after release.
%But even after possible changes to the software in the context of updating and maintenance work,
However, safety constraints can be applied again and assure the adherence of all properties. Safety constraints thus are also suitable for sustaining consistency during the different phases of software life. In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that. The more constraints are defined the the better can be the reliability.
But as already mentioned before, it might be difficult for developers to find all reasonable %safety
constraints by hand. % which are needed for full safety covery.
In this case it can happen that a prior correct software version becomes faulty over time because of minor adjustments once in a while which are not covered by constraints.
%To help prevent this to happen

As a solution to this problem, the concept of automated constraint generation for state machines is proposed. Once the initial implementation of a program is finished, the behaviour of this state machine can be analyzed. In particular heuristics can be elaborated which determine significant parts of the state machine automatically and put them into constraint representation. These constraints are of course valid on the current program.
On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg.\ if a computed constraint does not make sense at all.
On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.

@@

%All associated states which match these properties can be combined to a 
In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined. Such a group of states of a state machine is named subgraph from now on. It is defined as follows:
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

@@

These five constraints apply to each subgraph matching the conditions mentioned above. Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state. Formula~\ref{eq:second} ensures that there will always at least one $\gamma$-state be eventually visited after each visit of an $\alpha$-state. The quite similar but stricter formula~\ref{eq:third} additionally demands a visit of at least one $\gamma$-state before the $\beta$-state can be reached. Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states. The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

>>>>

Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications. In order to discover whether a program meets certain safety requirements, its behaviour is checked by the constraints. This happens mostly as a last check before software gets released. 
But regarding updating and maintenace work, often further changes are made to the software even after release.
%But even after possible changes to the software in the context of updating and maintenance work,
However, safety constraints can be applied again and assure the adherence of all properties. Safety constraints thus are also suitable for sustaining consistency during the different phases of software life. In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that. The more constraints are defined the better can be the reliability.
But as already mentioned before, it might be difficult for developers to find all reasonable %safety
constraints by hand. % which are needed for full safety covery.
In this case it can happen that a prior correct software version becomes faulty over time because of minor adjustments once in a while which are not covered by constraints.
%To help prevent this to happen

As a solution to this problem, the concept of automated constraint generation for state machines is proposed. Once the initial implementation of a program is finished, the behaviour of this state machine can be analyzed. In particular, heuristics can be elaborated which determine significant parts of the state machine automatically and put them into constraint representation. These constraints are of course valid on the current program.
On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg.\ if a computed constraint does not make sense at all.
On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.

@@

%All associated states which match these properties can be combined to a 
In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\gamma$-states and the one $\beta$-state have to be determined. Such a group of states of a state machine is named subgraph from now on. It is defined as follows:
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

@@

These five constraints apply to each subgraph matching the conditions mentioned above. Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state. Formula~\ref{eq:second} ensures that there will always at least one $\gamma$-state be eventually visited after each visit of an $\alpha$-state. The quite similar but stricter formula~\ref{eq:third} additionally demands a visit of at least one $\gamma$-state before the $\beta$-state can be reached. Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can not be visited between $\alpha$- and $\beta$-states. The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

### 2012-11-15 thesis/tex/6_theltlcreatorprototype.tex
final!

For the checking of constraints as well as for the constraint generation a model is needed which represents the state machine program. Three classes are responsible for holding the state machine program with all its necessary information: The class \emph{Fsm} represents the finite state machine and lists all contained states. It is called finite state machine, because the number of states is finite which applies to the domain of this work. \emph{State} represents the states of the machine and holds all transitions from and to the respective state. Transitions are represented by the class \emph{Transition} which specifies the source and a target state. Figure~\ref{fig:fsm_class} shows all three class definitions for the state machine model.

@@

As proposed in the requirements for the visual language in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: A vertical read direction and the horizontal direction for variation in time. 
Swing's mechanism for layouting components is used for the positioning of operators. After determining the aspired sizes of child components which represent sub operators, reference lines which have been already demonstrated with figure~\ref{fig:referencelines} in section~\ref{sec:operatorconstraintformalism} are computed mathematically. Based on this information operators are sized and positioned correctly.

@@

The overall implemented way of subgraph finding is illustrated in pseudo code with algorithm~\ref{alg:subgraphfinding}.
\begin{algorithm}
\caption{Finding of all subgraphs.}
\label{alg:subgraphfinding}
\begin{algorithmic}[1]
\STATE $subgraphs$: List
\FORALL{states as $S$}
  \STATE $start \leftarrow S$
	\STATE doLabelling($start$);
	\IF{state $E$ exists with label($E$)$=$label($S$)}
		\STATE $end \leftarrow E$
	\ELSE
		\STATE $end \leftarrow start$
	\ENDIF
	\IF{noCycles($start$, $end$) $and$ noExtTransitions($start$, $end$)}
		\STATE $subgraphs$.add($start$, $end$);
	\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}
Line 1 defines a variable $subgraphs$ which will contain all results of the algorithm after execution.
In line 4 the labelling process described above is executed by using one particular start state. Line 5 checks whether there is a state having the same label as the start state. If so, it is assumed as potential end state (line 6). Otherwise, the entire state machine is assumed to be a subgraph and the start state is defined as end state as well (line 8). Subsequent checks in line 9 finally determine whether it can be considered as proper subgraph.
There are no cycles allowed between start and end state since they are forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
Furthermore also foreign transitions are prohibited: Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph. If all checks are successful, start and end states get added to the result list $subgraphs$ in line 11. This procedure is applied to all states of the state machine by the surrounding foreach loop (line 2). In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.

@@

Both features, the editor for the visual language presented in section~\ref{sec:operatorconstraintformalism} and the automated constraint generation in chapter~\ref{chap:automatedgenerationofsafetyconstraints}, were assembled in one tool, the \emph{LTLCreator}.
Figure~\ref{fig:LTLCreator} shows the assembled parts of LTLCreator: A dashboard is provided where constraints can be edited visually by the developer and result displays indicate the validation results of constraints graphically. For validating the constraints an external model checker is linked to \emph{LTLCreator}. This component does not have an UI as well as the constraint generator, which is also part of \emph{LTLCreator}. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.65\linewidth]{LTLCreator}
  \caption{Integration of LTLCreator.}
  \label{fig:LTLCreator}
\end{figure}
The \emph{LTLCreator} itself does not provide functionalities for editing state machines although a state machine model is needed for the visual language as well as for constraint generation. Therefore the standalone version of \emph{LTLCreator} uses a hard coded state machine as model which represents an example program for test purpose.

@@

A main requirement to the \emph{LTLCreator} tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs. For this a \emph{ConstraintValidator} class has been defined which provides a method \emph{validate(AbstractOperator)} for proving constraints.
For the constraint validation itself a particular model checker can be specified. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used. It is an open source project and has been designed to be an open architecture for model checking. NuSMV does not have a java API but it can be externally accessed over command line. For this both the state machine program and the LTL formulas have to be translated to an input file which has to match a NuSMV specific format. For LTL formulas NuSMV uses the char ``G'' (for globally) instead of $\Box$ and the char ``F'' (for future) instead of $\Diamond$. An example of this input language is given in listing~\ref{lst:nusmv}.

@@

LTLSPEC G (state = 5 -> (F (state = 8 | state = 11)))
\end{lstlisting}
A state machine is encoded in the variable \emph{state}, and its value represents the active state. Thus, each state is identified by a number. The state machine described in the example of listing~\ref{lst:nusmv} contains eleven states, this is determined by the number range for \emph{state} from one to eleven. All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions. The initial state can be determined and a case distinction regulates the next possible values, i.e. transitions to reachable states. If there are more than one outgoing transition, all target state identifiers have to be surrounded by brackets. The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
The last line contains the constraint formula we want to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.

@@

As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the \emph{LTLCreator} module to revalidate all existing constraints. This demands a concept of change notification management which is responsible for receiving all changes and making this information available for the \emph{LTLCreator} component. As a solution, all parts of the code where model changes are made should be extended with short code snippets which send notifications.
However, this is not as easy to realise in Robostudio since some of the relevant code was generated or even imported as external jars which makes code manipulation impossible.

A second challenge is the current data handling policy in Robostudio. Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than for syntax checks. Thus syntax errors are not detected or even ignored by static analysis or input restrictions. In fact, such features haven't been requirements for Robostudio. During integration of \emph{LTLCreator} into Robostudio problems came up like transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistent data retrieved from Robostudio.

>>>>

For the checking of constraints as well as for the constraint generation a model is needed which represents the state machine program. Three classes are responsible for holding the state machine program with all its necessary information: The class \emph{Fsm} represents the finite state machine and lists all contained states. It is called finite state machine, because the number of states is finite which applies to the domain of this work. \emph{State} represents the states of the machine and holds a list of all transitions from or to itself. Transitions are represented by the class \emph{Transition} which specifies the source and a target state. Figure~\ref{fig:fsm_class} shows all three class definitions for the state machine model.

@@

As proposed in the design decisions for the visual language in section~\ref{sec:designdecisions}, constraints have particular meanings for the two dimensions: A vertical read direction and the horizontal direction for variation in time. 
Swing's mechanism for layouting components is used for the positioning of operators. After determining the aspired sizes of child components which represent sub operators, reference lines which have been already demonstrated with figure~\ref{fig:referencelines} in section~\ref{sec:designdecisions} are computed mathematically. Based on this information operators are sized and positioned correctly.

@@

The overall implemented way of subgraph finding is illustrated in pseudo code with algorithm~\ref{alg:subgraphfinding}.
\begin{algorithm}
\caption{Finding of all subgraphs.}
\label{alg:subgraphfinding}
\begin{algorithmic}[1]
\STATE $subgraphs$: List
\FORALL{states as $S$}
  \STATE $start \leftarrow S$
	\STATE doLabelling($start$);
	\IF{state $E$ exists with label($E$)$=$label($S$)}
		\STATE $end \leftarrow E$
	\ELSE
		\STATE $end \leftarrow start$
	\ENDIF
	\IF{noCycles($start$, $end$) $and$ noExtTransitions($start$, $end$)}
		\STATE $subgraphs$.add($start$, $end$);
	\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}
Line 1 defines a variable $subgraphs$ which will contain all results of the algorithm after execution.
In line 4 the labelling process described above is executed by using one particular start state. Line 5 checks whether there is a state having the same label as the start state. If so, it is assumed as potential end state (line 6). Otherwise, the entire state machine is assumed to be a subgraph and the start state is defined as end state as well (line 8). Subsequent checks in line 9 finally determine whether it can be considered as proper subgraph.
There are no cycles allowed between start and end state since they are forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
Furthermore also foreign transitions are prohibited: Transitions coming from a state outside the subgraph are allowed to lead to the start state or the end state but not to other states of the subgraph. If all checks are successful, start and end states get added to the result list $subgraphs$ in line 11. This procedure is applied to all states of the state machine by the surrounding foreach loop (line 2). In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.

@@

Both features, the editor for the visual language presented in section~\ref{sec:operatorconstraintformalism} and the automated constraint generation in chapter~\ref{chap:automatedgenerationofsafetyconstraints}, were assembled in one tool, the \emph{LTLCreator}.
Figure~\ref{fig:LTLCreator} shows the assembled parts of \emph{LTLCreator}: A dashboard is provided where constraints can be edited visually by the developer and result displays indicate the validation results of constraints graphically. For validating the constraints an external model checker is linked to \emph{LTLCreator}. This component does not have an UI as well as the constraint generator, which is also part of \emph{LTLCreator}. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.65\linewidth]{LTLCreator}
  \caption{Integration of LTLCreator.}
  \label{fig:LTLCreator}
\end{figure}
The \emph{LTLCreator} itself does not provide functionalities for editing state machines although a state machine model is needed for the visual language as well as for constraint generation. Therefore the standalone version of \emph{LTLCreator} uses a hard coded state machine as model which represents an example program for test purpose.

@@

A main requirement to the \emph{LTLCreator} tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs. For this a \emph{ConstraintValidator} class has been defined which provides a method \emph{validate(AbstractOperator)} for proving constraints.
For the constraint validation itself a particular model checker can be specified. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used. It is an open source project and has been designed to be an open architecture for model checking. NuSMV does not have a Java API but it can be externally accessed over command line. For this both the state machine program and the LTL formulas have to be translated to an input file which has to match a NuSMV specific format. For LTL formulas NuSMV uses the char ``G'' (for globally) instead of $\Box$ and the char ``F'' (for future) instead of $\Diamond$. An example of this input language is given in listing~\ref{lst:nusmv}.

@@

LTLSPEC G (state = 5 -> (F (state = 8 | state = 11)))
\end{lstlisting}
A state machine is encoded in the variable \emph{state}, and its value represents the active state. Thus, each state is identified by a number. The state machine described in the example of listing~\ref{lst:nusmv} contains eleven states, this is determined by the number range for \emph{state} from one to eleven. All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions. The initial state can be determined and a case distinction regulates the next possible values, i.e. transitions to reachable states. If there are more than one outgoing transition, all target state identifiers have to be surrounded by brackets. The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
The last line contains the constraint formula which has to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.

@@

As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the \emph{LTLCreator} module to revalidate all existing constraints. This demands a concept of change notification management which is responsible for receiving all changes and making this information available for the \emph{LTLCreator} component. As a solution, all parts of the code where model changes are made should be extended with short code snippets which send notifications.
However, this is not as easy to realize in Robostudio since some of the relevant code was generated or even imported as external jars which makes code manipulation impossible.

A second challenge is the current data handling policy in Robostudio. Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than for syntax checks. Thus, syntax errors are not detected or even ignored by methods like static analysis or input restrictions. In fact, such features have not been requirements for Robostudio. During integration of \emph{LTLCreator} into Robostudio problems came up when loading already existing RBDL programs of the healthbot project: Transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete and correct state machine as input and does not know how to cope with inconsistent data retrieved from Robostudio.
